Making Sequential Programs Deterministic

### 2019-04-05 java-design.tex
Edits in defaulting section

We have implemented our type system for Java, in a tool named \theDeterminismChecker.
The implementation consists of 1200 lines of Java
code, plus 1800 lines of tests (all line measurements are non-comment,
non-blank lines), a manual, examples, etc.  It is publicly available at
\myurl{URL for \theDeterminismChecker}.

\TheDeterminismChecker requires Java version 8.

@@

\TheDeterminismChecker supports the type qualifiers \<@NonDet>,
\<@OrderNonDet>, and \<@Det> (and others described below).

@@

\TheDeterminismChecker applies a default qualifier at each unannotated type
use (except uses of type variables, which already stand for a type that was
defaulted at the instantiation site).
This does not change the expressivity of the type system; it merely makes
the system more convenient to use.

In \theDeterminismChecker, we chose \<PolyDet> as the default for all 
method parameters and return types including constructors.
The rationale for this design choice was to not make \theDeterminismChecker too restrictive by making everything \<Det>.  This implies that
\theDeterminismChecker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.

@@

void bar(@NonDet int random, @Det int value) {
    @NonDet int x = foo(random);      // No error.
    @Det int y = foo(value);          // No error.
    System.out.println(foo(random));  // Error.
}
\end{verbatim}
In the example above, the method \codeid{foo} is annotated with the \<PolyDet> type qualifier which allows
the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. \TheDeterminismChecker flags an error 
only when the returned value is used in a way that breaks any deterministic specification (Printing a \<NonDet> value
breaks our determinism guarantees).

\<@Det> equality is with respect to \<.equals()>.

\todo{
The default annotation for most unannotated types is \<@Det>.

\begin{itemize}
\item
  Formal parameter and return types default to \<@PolyDet>.
\item
  If a formal parameter/return type is an array, its component type defaults to
  \<@PolyDet>.
\item
  As an exception, the return type of a method with no unannotated or
  \<@PolyDet> formal parameters (including the receiver) defaults to
  \<@Det>.
\end{itemize}

These choices permit all possible arguments to an unannotated method.
}

>>>>

We have implemented our type system for Java, in a tool named \theDeterminismChecker.
The implementation consists of 1200 lines of Java
code, plus 1800 lines of tests (all line measurements are non-comment,
non-blank lines), a manual, examples, etc.
\TheDeterminismChecker requires Java version 8.
It is publicly available at
\myurl{URL for \theDeterminismChecker}.

@@

\TheDeterminismChecker supports the type qualifiers \<@NonDet>,
\<@OrderNonDet>, and \<@Det> (and others described below).
The meaning of \<@Det> is with respect to value
equality, not reference equality; that is,
values on different executions are the same with respect to \<.equals()>,
not \<==>.

@@

\TheDeterminismChecker applies a default qualifier at each unannotated type
use (except uses of type variables, which already stand for a type that was
defaulted at the instantiation site).
This does not change the expressivity of the type system; it merely makes
the system more convenient to use by reducing programmer effort and code clutter.

The default annotation for most unannotated types is \<@Det>, because
programmers generally expect their programs to behave the same when re-run
on the same inputs.

Formal parameter and return types default to \<@PolyDet>.  That is, a
programmer-written method

\begin{Verbatim}
  int plus(int a, int b) { ... }
\end{Verbatim}

\noindent
is treated as if the programmer had written

\begin{Verbatim}
  @PolyDet int plus(@PolyDet int a, @PolyDet int b) { ... }
\end{Verbatim}

\noindent
and its function type is

$\forallt{\kappa}  \kappa\ \<int> \times \kappa\ \<int> \rightarrow \kappa\ \<int>$

\noindent
This choice type-checks if the method body does not make calls to any
interfaces that require \<@Det> arguments or produce \<@NonDet> results.

\begin{itemize}
\item
  If a formal parameter/return type is an array, its component type defaults to
  \<@PolyDet>.
\item
  As an exception, the return type of a method with no unannotated or
  \<@PolyDet> formal parameters (including the receiver) defaults to
  \<@Det>.
\end{itemize}

These choices permit all possible arguments to an unannotated method.

In \theDeterminismChecker, we chose \<PolyDet> as the default for all 
method parameters and return types including constructors.
The rationale for this design choice was to not make \theDeterminismChecker too restrictive by making everything \<Det>.

@@

void bar(@NonDet int random, @Det int value) {
    @NonDet int x = foo(random);      // no error
    @Det int y = foo(value);          // no error
    System.out.println(foo(random));  // error
}
\end{verbatim}
In the example above, the method \codeid{foo} is annotated with the \<PolyDet> type qualifier which allows
the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. \TheDeterminismChecker flags an error 
only when the returned value is used in a way that breaks any deterministic specification (Printing a \<NonDet> value
breaks our determinism guarantees).
